Compiling Python Modules to Native Parallel Modules Using Pythran and OpenMP Annotations

### 2013-09-04 doc/papers/sc2013/paper.tex
SC2013: fix affilitation

\author{
    \IEEEauthorblockN{Serge Guelton\IEEEauthorrefmark{1}, \IEEEauthorrefmark{2}}
    \IEEEauthorblockN{Pierrick Brunet\IEEEauthorrefmark{2}}
    \IEEEauthorblockN{Mehdi Amini\IEEEauthorrefmark{3}}
    \newline
    \IEEEauthorblockA{\newline\IEEEauthorrefmark{1}
        {\'E}cole Normale Sup{\'e}rieure, D\'{e}partement d'Informatique, Paris, France}
    \IEEEauthorblockA{\newline\IEEEauthorrefmark{2}
    T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France}
    \IEEEauthorblockA{\newline\IEEEauthorrefmark{3}
    Silkan, Meudon la For\^et, France}
}

@@

    HPC users traditionally rely on low-level, compiled language such as C or
    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
    common situation to have HPC application written in a high-level language
    such as Python, calling native routines for computation-intensive tasks. To
    improve development speed and reduce maintenance costs, several solutions
    such as Cython, Numba, Parakeet or Pythran, have appeared to make it
    possible to automatically or semi-automatically turn Python functions into
    native ones.

    One of the key point to match the performance of native applications is the
    ability to write parallel applications. This paper studies the addition of
    OpenMP directives, a popular model to describe the parallelism of
    C/C++/FORTRAN applications, to Pythran, an automatic compiler from a subset
    of Python to C++. It shows that scientific Python applications annotated
    with OpenMP directives can be turned by an automatic compiler into native
    applications that run within the same order of magnitude than
    manually-written ones.

\end{abstract}

@@

Since its birth in December 1989, the Python language~\cite{rossum97} has proved
to be useful in various domains, ranging from system administration to web
services, thanks to its dynamicity, expressiveness, rich ecosystem and
battery-included standard library. It is also getting widely used in scientific
computing~\cite{Oliphant2007}, mainly thanks to the SciPy~\cite{scipy} project
that provides a set of scientific and numeric tools, ranging from linear algebra
to ordinary differential equation solvers or generic algorithms.

It is now heading toward High Performance Computing(HPC), either as a glue
language used to bind several native libraries together, or to write whole
applications in Pure Python. However, the prohibitive performance overhead of
the language implied by its interpreted and highly dynamic nature prevents its
usage for the most performance - critical code sections, where native code is
generally used. The SciPy modules partly overcome this issue through the usage
of low-level routines written in C or Fortran and encapsulated in Python
\emph{native} modules. Moreover, its core data structure, the multi-dimensional
array~\cite{numpyarray2011}, has been designed so that the underlying data are
available to both native modules and the Python interpreter without conversion
cost. Yet when a new function not already part of an existing module is
required, one has to write it in C or FORTRAN.

To enable the user to write native functions without having to write the glue
code that turns Python object into native structures back and forth, the SciPy
package provides the \texttt{weave} module that makes it possible to bundle C
code snippet into Python code and compile then load them at runtime. This hybrid
approach, based on a mix of interpreted and native code, is getting widespread
in the Python landscape. It is also the recommended approach to write
application that use fine grain parallelism, has Python parallelism support
targets coarse-grain parallelism.

This paper presents the addition of OpenMP, a popular standard to turn
sequential C, C++ and FORTRAN application into parallel applications, to the
Pythran~\cite{pythran2013} compiler, a compiler that turn Python modules into
C++ meta-programs.  Section~\ref{sec:python-parallel} gives an overview of
existing approaches to compile Python functions or modules and shows a critical
lack of parallelism support, and emphasises on the need for a
backward-compatible approach.  Section~\ref{sec:python-static} presents the
Pythran compiler and the underlying runtime. Section~\ref{sec:python-openmp}
studies the validity of using OpenMP directives within the Python language in
order to add fine-grain parallelism support to Python in the context of Pythran.
The Pythran compiler is benchmarked on several scientific kernels, compared
to Cython, Numba and Parakeet in Section~\ref{sec:validation}.

@@

In the many cores area, turning to parallelism to balance the performance
limitations of scripting languages, as described in~\cite{choy05} in the context
of the MATLAB language, or in~\cite{mals07} in the context of the R language,
seems legitimate. In the context of Python, most approaches have focused on
fork-based parallelism.

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that can
communicate through Inter Process Communication(IPC), using Python built-in
object serialization. This approach is only viable for computation intensive
application, since the communication and synchronization overheads are much
greater than a light-weight processes approach.

The standard \texttt{threading} module also makes it possible to start
several light-weight threads within the same interpreter, but this lower-level approach is not
applicable to HPC, because of a specificity of CPython, the \emph{Global
Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
is active at a time in the interpreter. While it makes it possible to have
cooperative threads, say for a GUI, it does not take advantage of multiple
cores.  However, there are two notable exceptions: the GIL is released on I/O,
and the GIL does not prevent the use of threads inside native modules, where the
user has full control.

@@

from multiprocessing import Pool
def multi_buffon(darts):
  p = Pool(4)
  return sum(p.map(buffon, [darts//4]*4))/4
\end{lstlisting}

@@

It is now common for scripting language to have C bindings. To take advantage of
compiled code, and to overcome the GIL limitations, Python developers must write
parallel C/C++ functions and the associated boilerplate based on the Python C
API~\cite{pythoncapi}. Tools have been developed to relieve the user from
writing the encapsulating code, notably SWIG~\cite{swig2003} that relies on an
enhanced interface specification, or
\texttt{boost::python}~\cite{boostpython2007} that relies on C++ facilities to
guide translation.

An opposite approach consists in using the host language ---herein Python--- to
describe both parts of the system, and to let an automated tool perform the
translation to native code of a part of the application, generally the
computation-intensive one where parallelism has been expressed in some ways.
Thus, developers not familiar with lower level languages or not eager to invest
the additional development time can still benefit from a fair performance boost.
This approach is the subject of many studies that can be classified according to
their compatibility with the host language.

@@

Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing informations, calls to native functions from third party
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variable to improve performance. Without type annotation, the performance
improvement is not terrific, but given enough type annotation, Cython can
generates code that runs as fast as its C equivalent, while maintaining a syntax
close to Python.

@@

The main issue of these approaches is that they imply to modify, in a
backward-incompatible way, the original code. They require to learn a new
dialect, and the long-term preservation of this investment is not ensured.
Moreover, there is no falling back if the code were to be deployed in an
environment where the parallelizing tool/module is not available. For instance,
once a code has been ported to PyCUDA and transformed into CUDA code embedded in
Python strings, there is no way back and the developer has to maintain two
versions of the algorithm.

@@

The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
compiler uses additional type information to generate sequential LLVM bytecode.
Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
Python to LLVM bytecode translation, but limits its scope to the numpy package,
only supporting a small subset of the Python language. Additionally, it supports
implicit parallelism using an implicit mapping between numpy functions and a set
of parallel primitives including \texttt{map}s \texttt{scan}s and
\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
not suffer from the backward-incompatibility issues. When meeting an unsupported
construct, numba falls back to Python C-API calls, while parakeet raises an
exception.

@@

This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with the Python subset supported by the Pythran compiler to make it possible to
write fine-grain parallel applications in Python, while being fully backward
compatible with both the Python language and the sequential algorithm. It means
that:

>>>>

\author{
    \IEEEauthorblockN{Serge Guelton\IEEEauthorrefmark{1}, \IEEEauthorrefmark{2}}
    \IEEEauthorblockN{Pierrick Brunet\IEEEauthorrefmark{2}}
    \IEEEauthorblockN{Mehdi Amini\IEEEauthorrefmark{3}}
    \newline
    \IEEEauthorblockA{\newline\IEEEauthorrefmark{1}
        {\'E}cole Normale Sup{\'e}rieure, D\'{e}partement d'Informatique, Paris, France}
    \IEEEauthorblockA{\newline\IEEEauthorrefmark{2}
    T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France}
    \IEEEauthorblockA{\newline\IEEEauthorrefmark{3}
    SILKAN Inc., Los-Altos, USA}
}

@@

    HPC users traditionally rely on low-level, compiled language such as C or
    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
    common situation to have HPC application written in a high-level language
    such as Python, calling native routines for computate-intensive tasks. To
    improve development speed and reduce maintenance costs, using a higher-level
    language like Python seems attractive. While it is usually associated with
    low performance, several solutions such as Cython, Numba, Parakeet or 
    Pythran offer to automatically or semi-automatically turn Python functions 
    into native ones.

    One of the key points required to match the performance of native 
    applications is the ability to write parallel applications. This paper
    studies the addition of OpenMP directives, a popular model to describe
    parallelism in C/C++/FORTRAN applications, to Pythran, an automatic compiler     
    from a subset of Python to C++. It shows that scientific Python applications
    annotated with OpenMP directives can be turned by an automatic compiler into 
    native applications that run within the same order of magnitude than
    manually-written ones.
\end{abstract}

@@

Since its birth in December 1989, the Python language~\cite{rossum97} has proved
to be useful in various domains, ranging from system administration to web
services, thanks to its dynamicity, expressiveness, rich ecosystem and
% Mehdi : battery-included ??????
battery-included standard library. It is also widely used in scientific
computing~\cite{Oliphant2007}, thanks to the SciPy~\cite{scipy} project
which provides a set of scientific and numeric tools, ranging from linear algebra to ordinary differential equation solvers or generic algorithms.

It is now heading toward High Performance Computing (HPC), either as a glue
language used to bind several native libraries together, or to support whole
applications in Pure Python. However, the prohibitive overhead of the language
implied by its interpreted and highly dynamic nature prevents its usage for the
most performance-critical code sections, where native code is generally used.
The SciPy modules partly overcome this issue through the usage of low-level 
routines written in C or Fortran and encapsulated in Python \emph{native} 
modules. Moreover, its core data structure, the multi-dimensional
array~\cite{numpyarray2011}, has been designed so that the underlying data are
available to both native modules and the Python interpreter without conversion
cost. Yet when a new function not already part of an existing module is
required, one has to write it in C or FORTRAN.

To allow the user to write native functions without having to write the boilerplate glue code that turns Python object into native structures back and
forth, the SciPy package provides the \texttt{weave} module. It makes
possible to bundle C code snippets into Python code. They are then compiled and 
loaded at runtime. This hybrid approach, based on a mix of interpreted and
native code, is getting widespread in the Python landscape. It is also the
recommended approach to write applications that make use of fine grain
parallelism, as Python parallelism support is only suitable for coarse-grain
parallelism.

This paper presents the addition of OpenMP, a popular standard to turn
sequential C, C++ and FORTRAN application into parallel applications, to the
Pythran~\cite{pythran2013} compiler, a compiler that turn Python modules into
C++ meta-programs.  Section~\ref{sec:python-parallel} gives an overview of
existing approaches to compile Python functions or modules and shows a critical
lack of parallelism support. It also emphasises on the need for a
backward-compatible approach. Section~\ref{sec:python-static} presents the
Pythran compiler and the underlying runtime. Section~\ref{sec:python-openmp}
studies the validity of using OpenMP directives within the Python language in
% I would dispute the fact that OpenMP is "fine-grain", we're not talking about % vectorization here.
order to add fine-grain parallelism support to Python in the context of Pythran.
Finally, the Pythran compiler is benchmarked on several scientific kernels and
compared to Cython, Numba and Parakeet in Section~\ref{sec:validation}.

@@

% Mehdi area => era, Serge: validate?
% I also wonder the rational of using parallelism when performance are bad 
% because language is bad…
In the many-cores era, it is mandatory to exhibit parallelism to balance the 
performance limitations of scripting languages, as described in~\cite{choy05}
in the context of the MATLAB language, or in~\cite{mals07} in the context of 
the R language. In the context of Python, most approaches have focused on
fork-based parallelism.

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that
communicate through Inter Process Communication(IPC), using Python built-in
object serialization. This approach is only viable for computation intensive
independent tasks, since the communication and synchronization overheads are 
much greater than a light-weight threaded approach.

The standard \texttt{threading} module also makes possible to start several
light-weight threads within the same interpreter, but this lower-level approach
is not applicable to HPC, because of a specificity of CPython, the \emph{Global
Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
is active at a time in the interpreter. While it enables the possibility of
cooperative threads, say for a GUI, it does not take advantage of multiple
cores. However, there are two notable exceptions: the GIL is released on I/O,
and the GIL does not prevent the use of threads inside native modules, where the
user has full control.

@@

from multiprocessing import Pool
def multi_buffon(darts):
  n = 4
  p = Pool(n)
  return sum(p.map(buffon, [darts//n]*n))/n
\end{lstlisting}
% Mehdi: I don't get the [darts//4]*4 at the last line of the listing

@@

It is now common for scripting language to have C bindings. To take advantage of
compiled code, and to overcome the GIL limitations, Python developers must write
parallel C/C++ functions and the associated boilerplate based on the Python C
API~\cite{pythoncapi}. Tools have been developed to relieve the user from this
cumbersome task, notably SWIG~\cite{swig2003} that relies on an interface specification supplied by the programmer to generate the glue, or 
\texttt{boost::python}~\cite{boostpython2007} that relies on C++ templates and 
type overloading facilities to guide translation.

An opposite approach consists in using the host language ---herein Python--- to
describe both parts of the system, i.e.\ the hybrid and the native. An automated 
tool performs the translation to native code of a specific part of the 
application, generally the computation-intensive one where parallelism has been
expressed in some ways.
Therefore, developers not familiar with lower level languages or not eager to 
invest the additional development time can still benefit from a fair performance
boost.
This approach is the subject of many studies that can be classified according to
their compatibility with the host language.

@@

Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing informations, calls to native functions from third party
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variables to improve performance. Without type annotations, the 
performance improvement is not terrific, but given enough type annotation, Cython can generates code that runs as fast as its C equivalent, while 
maintaining a syntax close to Python.

@@

The main drawback of these approaches is that they require to modify in a
backward-incompatible way the original code. They require to learn a new
dialect, and the long-term preservation of this investment is not ensured.
Moreover, there is no fall back if the code were to be deployed in an
environment where the parallelizing tool/module is not available. For instance,
once a code has been ported to PyCUDA and transformed into CUDA code embedded in
Python strings, there is no way back and the developer has to maintain two
versions of the algorithm.

@@

The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
compiler uses additional type information to generate sequential LLVM bytecode.
Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
Python to LLVM bytecode translation, but limits its scope to the numpy package,
only supporting a small subset of the Python language. Additionally, it supports
implicit parallelism using an implicit mapping between numpy functions and a set
of parallel primitives including \texttt{map}s \texttt{scan}s and
\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
not suffer from backward-incompatibility issues. When meeting an unsupported
construct, numba falls back to Python C-API calls, while parakeet raises an
exception.

@@

%FIXME: why not cite OpenMP 4.0 (aka the latest)?
This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with the Python subset supported by the Pythran compiler to make it possible to
write fine-grain parallel applications in Python, while being fully backward
compatible with both the Python language and the sequential algorithm. It 
implies that:
